---
title: "Brighton & Hove: Defibrillator Provision vs. Daytime Demand"
author: "Adam Dennett"
format: html
---

## Introduction

This analysis compares the current provision of **24/7 publicly accessible** Automated External Defibrillators (AEDs) in Brighton & Hove against estimated daytime population demand. Using Census 2021 workday population density at Lower Super Output Area (LSOA) level as a proxy for where people are during the day, we identify areas with high demand but low defibrillator coverage — and rank them by priority for new AED placement.

**Data sources:**

- AED locations from [The Circuit](https://www.defibfinder.uk/) (February 2026 extract)
- Census 2021 Workday Population Density (table WD001) from [Nomis](https://www.nomisweb.co.uk/sources/census_2021_wd)
- LSOA 2021 boundaries from the [ONS Open Geography Portal](https://geoportal.statistics.gov.uk/)

## Load Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

library(readxl)
library(sf)
library(dplyr)
library(tidyr)
library(readr)
library(leaflet)
library(tmap)
library(classInt)
library(htmltools)
library(knitr)
library(kableExtra)
```

## Load and Filter Defibrillator Data

```{r}
#| label: load-defib
#| message: false

# Load full dataset
defib <- read_excel("defibrillator_data.xlsx",
                     sheet = "data_extract_2026-02-02")

# Filter to Brighton & Hove
bh_defib <- defib %>%
  filter(grepl("Brighton", ladnm, ignore.case = TRUE),
         !is.na(lat), !is.na(long))

cat("Total AEDs in Brighton & Hove:", nrow(bh_defib), "\n")

# Summary by availability and access type
bh_defib %>%
  count(defibrillators_availability, defibrillators_access_type,
        name = "count") %>%
  arrange(desc(count)) %>%
  kable(col.names = c("Availability", "Access Type", "Count")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
#| label: filter-247-public

# Filter to 24/7 public access defibrillators only
bh_247_public <- bh_defib %>%
  filter(defibrillators_availability == "24/7 Access",
         defibrillators_access_type == "Public")

cat("24/7 Public AEDs in Brighton & Hove:", nrow(bh_247_public), "\n")

# Convert to spatial
bh_247_sf <- st_as_sf(bh_247_public, coords = c("long", "lat"), crs = 4326)

# Also create sf for ALL B&H defibrillators (for context)
bh_all_sf <- st_as_sf(bh_defib, coords = c("long", "lat"), crs = 4326)
```

## Download External Data

### LSOA 2021 Boundaries

```{r}
#| label: download-boundaries
#| message: false

# Download Brighton & Hove LSOA boundaries from ONS ArcGIS FeatureServer
# Using the BSC (Super Generalised Clipped) version for web mapping
api_url <- paste0(
  "https://services1.arcgis.com/ESMARspQHYMw9BZ9/arcgis/rest/services/",
  "Lower_layer_Super_Output_Areas_December_2021_Boundaries_EW_BSC_V4/",
  "FeatureServer/0/query?",
  "where=LSOA21NM+LIKE+%27Brighton%25%27",
  "&outFields=LSOA21CD,LSOA21NM",
  "&returnGeometry=true",
  "&outSR=4326",
  "&f=geojson",
  "&resultRecordCount=200"
)

# Download to temp file and read
tmp_geojson <- tempfile(fileext = ".geojson")
download.file(api_url, tmp_geojson, quiet = TRUE, mode = "wb")
bh_lsoa <- st_read(tmp_geojson, quiet = TRUE)

cat("LSOA boundaries loaded:", nrow(bh_lsoa), "LSOAs\n")
```

### Census 2021 Workday Population Density (WD001)

```{r}
#| label: download-wd001
#| message: false

# Download WD001 from Nomis
tmp_zip <- tempfile(fileext = ".zip")
download.file("https://www.nomisweb.co.uk/output/census/2021/wd001.zip",
              tmp_zip, quiet = TRUE, mode = "wb")

# Extract and read the LSOA-level CSV
tmp_dir <- tempdir()
unzip(tmp_zip, exdir = tmp_dir)

wd001 <- read_csv(file.path(tmp_dir, "WD001_lsoa.csv"),
                   show_col_types = FALSE)

# Rename columns for easier use
wd001 <- wd001 %>%
  rename(
    lsoa21cd = `Lower layer Super Output Areas Code`,
    lsoa21nm = `Lower layer Super Output Areas Label`,
    workday_density = `Population Density`
  )

# Filter to Brighton & Hove LSOAs
bh_wd001 <- wd001 %>%
  filter(lsoa21cd %in% bh_lsoa$LSOA21CD)

cat("Workday density data for", nrow(bh_wd001), "Brighton & Hove LSOAs\n")
cat("Range:", round(min(bh_wd001$workday_density)),
    "to", round(max(bh_wd001$workday_density)),
    "persons per km²\n")
```

## Build Analysis Dataset

```{r}
#| label: build-dataset

# Count 24/7 public AEDs per LSOA
aed_counts <- bh_247_public %>%
  count(lsoa21, name = "aed_count")

# Join workday density to LSOA boundaries
bh_analysis <- bh_lsoa %>%
  left_join(bh_wd001, by = c("LSOA21CD" = "lsoa21cd")) %>%
  left_join(aed_counts, by = c("LSOA21CD" = "lsoa21")) %>%
  mutate(
    aed_count = replace_na(aed_count, 0),
    # Priority score: higher workday density with fewer AEDs = higher priority
    priority_score = workday_density / (aed_count + 1),
    priority_rank = rank(-priority_score, ties.method = "first")
  )

cat("LSOAs with zero 24/7 public AEDs:",
    sum(bh_analysis$aed_count == 0), "of", nrow(bh_analysis), "\n")
```

## Map 1: Workday Population Demand with AED Locations

This map shows the Census 2021 workday population density across Brighton & Hove LSOAs. Darker shading indicates higher concentrations of people during the working day. Red points show the locations of existing 24/7 publicly accessible defibrillators.

```{r}
#| label: map-demand
#| fig-width: 10
#| fig-height: 8

tmap_mode("plot")

tm_shape(bh_analysis) +
  tm_polygons(
    fill = "workday_density",
    fill.scale = tm_scale_continuous(
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Workday Pop.\nDensity\n(per km²)"
    ),
    col = "grey60",
    lwd = 0.5
  ) +
  tm_shape(bh_247_sf) +
  tm_symbols(
    fill = "red",
    col = "darkred",
    size = 0.3,
    shape = 21
  ) +
  tm_title("Workday Population Demand & 24/7 Public AED Locations") +
  tm_layout(
    frame = FALSE
  ) +
  tm_scalebar(position = c("left", "bottom"))
```

## Map 2: Interactive Demand & Provision Map

Zoom in to explore individual LSOAs and AED locations. Click on LSOAs for workday population data, or on red markers for AED details.

```{r}
#| label: map-interactive

# Create colour palette for workday density
pal_density <- colorNumeric("Blues", domain = bh_analysis$workday_density)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Choropleth layer
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_density(workday_density),
    fillOpacity = 0.6,
    color = "grey50",
    weight = 1,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Workday density:</em> ",
      format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count, "<br>",
      "<em>Priority rank:</em> ", priority_rank, " of ", nrow(bh_analysis)
    ),
    group = "Workday Density"
  ) %>%
  # AED point markers
  addCircleMarkers(
    data = bh_247_sf,
    radius = 5,
    color = "darkred",
    fillColor = "red",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_density,
    values = bh_analysis$workday_density,
    title = "Workday Density<br>(per km²)"
  ) %>%
  addLayersControl(
    overlayGroups = c("Workday Density", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

## Map 3: Coverage Gap — Priority Areas for New AEDs

The priority score is calculated as: **workday population density ÷ (number of existing 24/7 public AEDs + 1)**. A higher score indicates an LSOA with high daytime demand but few or no defibrillators — and therefore greater need for new provision.

LSOAs outlined in red are the **top 20 priority** areas.

```{r}
#| label: map-priority
#| fig-width: 10
#| fig-height: 8

# Flag top 20 priority LSOAs
bh_analysis <- bh_analysis %>%
  mutate(top20 = priority_rank <= 20)

tmap_mode("plot")

tm_shape(bh_analysis) +
  tm_polygons(
    fill = "priority_score",
    fill.scale = tm_scale_continuous(
      values = "brewer.yl_or_rd"
    ),
    fill.legend = tm_legend(
      title = "Priority\nScore"
    ),
    col = "grey60",
    lwd = 0.5
  ) +
  # Highlight top 20
  tm_shape(bh_analysis %>% filter(top20)) +
  tm_borders(col = "red", lwd = 2) +
  # Existing AEDs
  tm_shape(bh_247_sf) +
  tm_symbols(
    fill = "black",
    col = "white",
    size = 0.25,
    shape = 21
  ) +
  tm_title("Priority Areas for New 24/7 Public AEDs") +
  tm_layout(
    frame = FALSE
  ) +
  tm_scalebar(position = c("left", "bottom"))
```

## Priority Ranking Table

The table below ranks all Brighton & Hove LSOAs by priority for new 24/7 public defibrillator placement. LSOAs with high workday population density but few or no existing AEDs are ranked highest.

```{r}
#| label: priority-table

priority_table <- bh_analysis %>%
  st_drop_geometry() %>%
  select(
    Rank = priority_rank,
    LSOA = LSOA21NM,
    `Workday Density (per km²)` = workday_density,
    `24/7 Public AEDs` = aed_count,
    `Priority Score` = priority_score
  ) %>%
  arrange(Rank) %>%
  mutate(
    `Workday Density (per km²)` = format(round(`Workday Density (per km²)`),
                                          big.mark = ","),
    `Priority Score` = format(round(`Priority Score`), big.mark = ",")
  )

# Show top 20
priority_table %>%
  head(20) %>%
  kable(align = c("r", "l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

### Full Ranking

```{r}
#| label: full-table

priority_table %>%
  kable(align = c("r", "l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  scroll_box(height = "400px")
```

## Summary

```{r}
#| label: summary-stats
#| echo: false

n_total <- nrow(bh_analysis)
n_zero <- sum(bh_analysis$aed_count == 0)
n_aeds <- nrow(bh_247_sf)
top1 <- bh_analysis %>% filter(priority_rank == 1) %>% st_drop_geometry()
```

**Key findings:**

- Brighton & Hove has **`r n_total` LSOAs** and **`r n_aeds` 24/7 publicly accessible AEDs**
- **`r n_zero` LSOAs (`r round(100 * n_zero / n_total)`%)** have **no** 24/7 public AED
- The highest-priority LSOA for new provision is **`r top1$LSOA21NM`** with a workday population density of `r format(round(top1$workday_density), big.mark = ",")` persons per km² and `r top1$aed_count` existing 24/7 public AED(s)
- The priority ranking combines both demand (workday population density) and existing supply, meaning LSOAs that already have good coverage are ranked lower regardless of their density

---

*Data sourced from [The Circuit](https://www.defibfinder.uk/) and [ONS Census 2021](https://www.nomisweb.co.uk/sources/census_2021_wd). Is this an emergency? This website is used to locate defibrillators, but it is not intended for use in an emergency. If you require urgent medical assistance, call 999 now.*
