---
title: "Brighton & Hove: Defibrillator Provision vs. Daytime & Resident Population Demand"
author: "Adam Dennett"
format: html
execute:
  warning: false
  message: false
---

## Introduction

This analysis compares the current provision of **24/7 publicly accessible** Automated External Defibrillators (AEDs) in Brighton & Hove against two complementary measures of population demand:

1.  **Workday population density** (Census 2021, table WD001) — a proxy for where people are during the daytime
2.  **Resident population density** (Census 2021, table TS001) — the usual resident population, reflecting overnight and evening demand

By examining both perspectives, we can identify areas with high demand but low defibrillator coverage and understand how priority rankings shift depending on whether we consider daytime or residential populations.

**Data sources:**

-   AED locations from [The Circuit](https://www.defibfinder.uk/) (February 2026 extract)
-   Census 2021 Workday Population Density (table WD001) from [Nomis](https://www.nomisweb.co.uk/sources/census_2021_wd)
-   Census 2021 Usual Resident Population (table TS001) from [Nomis](https://www.nomisweb.co.uk/sources/census_2021_bulk)
-   LSOA 2021 boundaries (BGC V5, Generalised Clipped) from the [ONS Open Geography Portal](https://geoportal.statistics.gov.uk/)
-   Telephone box locations from [OpenStreetMap](https://www.openstreetmap.org/) via the Overpass API

## Load Libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

library(readxl)
library(sf)
library(dplyr)
library(tidyr)
library(readr)
library(leaflet)
library(tmap)
library(classInt)
library(htmltools)
library(knitr)
library(kableExtra)
library(jsonlite)
library(ggplot2)
```

## Load and Filter Defibrillator Data

```{r}
#| label: load-defib
#| message: false

# Load full dataset
defib <- read_excel("defibrillator_data.xlsx",
                     sheet = "data_extract_2026-02-02")

# Filter to Brighton & Hove
bh_defib <- defib %>%
  filter(grepl("Brighton", ladnm, ignore.case = TRUE),
         !is.na(lat), !is.na(long))

cat("Total AEDs in Brighton & Hove:", nrow(bh_defib), "\n")

# Summary by availability and access type
bh_defib %>%
  count(defibrillators_availability, defibrillators_access_type,
        name = "count") %>%
  arrange(desc(count)) %>%
  kable(col.names = c("Availability", "Access Type", "Count")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
#| label: filter-247-public

# Filter to 24/7 public access defibrillators only
bh_247_public <- bh_defib %>%
  filter(defibrillators_availability == "24/7 Access",
         defibrillators_access_type == "Public")

cat("24/7 Public AEDs in Brighton & Hove:", nrow(bh_247_public), "\n")

# Convert to spatial
bh_247_sf <- st_as_sf(bh_247_public, coords = c("long", "lat"), crs = 4326)

# Also create sf for ALL B&H defibrillators (for context)
bh_all_sf <- st_as_sf(bh_defib, coords = c("long", "lat"), crs = 4326)
```

## Download External Data

### LSOA 2021 Boundaries

```{r}
#| label: download-boundaries
#| message: false

# Download Brighton & Hove LSOA boundaries from ONS ArcGIS FeatureServer
# Using the BGC (Generalised Clipped) V5 version for higher-quality boundaries
api_url <- paste0(
  "https://services1.arcgis.com/ESMARspQHYMw9BZ9/arcgis/rest/services/",
  "Lower_layer_Super_Output_Areas_December_2021_Boundaries_EW_BGC_V5/",
  "FeatureServer/0/query?",
  "where=LSOA21NM+LIKE+%27Brighton%25%27",
  "&outFields=LSOA21CD,LSOA21NM",
  "&returnGeometry=true",
  "&outSR=4326",
  "&f=geojson",
  "&resultRecordCount=200"
)

# Download to temp file and read
tmp_geojson <- tempfile(fileext = ".geojson")
download.file(api_url, tmp_geojson, quiet = TRUE, mode = "wb")
bh_lsoa <- st_read(tmp_geojson, quiet = TRUE)

cat("LSOA boundaries loaded:", nrow(bh_lsoa), "LSOAs\n")
```

### Census 2021 Workday Population Density (WD001)

```{r}
#| label: download-wd001
#| message: false

# Download WD001 from Nomis
tmp_zip <- tempfile(fileext = ".zip")
download.file("https://www.nomisweb.co.uk/output/census/2021/wd001.zip",
              tmp_zip, quiet = TRUE, mode = "wb")

# Extract and read the LSOA-level CSV
tmp_dir <- tempdir()
unzip(tmp_zip, exdir = tmp_dir)

wd001 <- read_csv(file.path(tmp_dir, "WD001_lsoa.csv"),
                   show_col_types = FALSE)

# Rename columns for easier use
wd001 <- wd001 %>%
  rename(
    lsoa21cd = `Lower layer Super Output Areas Code`,
    lsoa21nm = `Lower layer Super Output Areas Label`,
    workday_density = `Population Density`
  )

# Filter to Brighton & Hove LSOAs
bh_wd001 <- wd001 %>%
  filter(lsoa21cd %in% bh_lsoa$LSOA21CD)

cat("Workday density data for", nrow(bh_wd001), "Brighton & Hove LSOAs\n")
cat("Range:", round(min(bh_wd001$workday_density)),
    "to", round(max(bh_wd001$workday_density)),
    "persons per km²\n")
```

### Census 2021 Usual Resident Population (TS001)

```{r}
#| label: download-ts001
#| message: false

# Download TS001 (usual resident population) from Nomis Census 2021 bulk data
tmp_zip_ts <- tempfile(fileext = ".zip")
download.file("https://www.nomisweb.co.uk/output/census/2021/census2021-ts001.zip",
              tmp_zip_ts, quiet = TRUE, mode = "wb")

# Extract to a dedicated subdirectory to avoid file conflicts
tmp_dir_ts <- file.path(tempdir(), "ts001")
dir.create(tmp_dir_ts, showWarnings = FALSE)
unzip(tmp_zip_ts, exdir = tmp_dir_ts)

# Find the LSOA-level CSV
ts_files <- list.files(tmp_dir_ts, pattern = "\\.csv$",
                        full.names = TRUE, recursive = TRUE)
ts001_file <- ts_files[grep("lsoa", ts_files, ignore.case = TRUE)][1]

ts001_raw <- read_csv(ts001_file, show_col_types = FALSE)

# Identify columns flexibly (Census 2021 bulk CSVs vary in naming)
code_col <- names(ts001_raw)[grep("geography.code|Areas.Code|LSOA.*Code",
                                   names(ts001_raw), ignore.case = TRUE)][1]
total_col <- names(ts001_raw)[grep("Total|Usual.resident",
                                    names(ts001_raw), ignore.case = TRUE)][1]

bh_ts001 <- ts001_raw %>%
  select(lsoa21cd = all_of(code_col),
         resident_pop = all_of(total_col)) %>%
  filter(lsoa21cd %in% bh_lsoa$LSOA21CD)

cat("Resident population data for", nrow(bh_ts001), "Brighton & Hove LSOAs\n")
cat("Total usual residents:", format(sum(bh_ts001$resident_pop), big.mark = ","), "\n")
```

### OpenStreetMap Telephone Box Locations

Telephone boxes are potential sites for new defibrillator installations — many across the UK have already been converted. We download current telephone box locations from [OpenStreetMap](https://www.openstreetmap.org/) via the Overpass API (`amenity=telephone`).

```{r}
#| label: download-phone-boxes

# Query OSM Overpass API for telephone boxes in Brighton & Hove
# Use a bounding box around Brighton & Hove to avoid URL encoding issues
# with area names. B&H bbox: south=50.79, west=-0.30, north=50.89, east=-0.04
overpass_query <- '[out:json][timeout:30];node[amenity=telephone](50.79,-0.30,50.89,-0.04);out body;'
overpass_url <- paste0(
  "https://overpass-api.de/api/interpreter?data=",
  URLencode(overpass_query, reserved = TRUE)
)

osm_raw <- fromJSON(overpass_url)

phone_boxes <- osm_raw$elements %>%
  select(id, lat, lon) %>%
  filter(!is.na(lat), !is.na(lon))

phone_sf <- st_as_sf(phone_boxes, coords = c("lon", "lat"), crs = 4326)

cat("Telephone boxes loaded from OSM:", nrow(phone_sf), "\n")
```

## Build Analysis Dataset

```{r}
#| label: build-dataset

# Count 24/7 public AEDs per LSOA
aed_counts <- bh_247_public %>%
  count(lsoa21, name = "aed_count")

# Join workday density to LSOA boundaries
bh_analysis <- bh_lsoa %>%
  left_join(bh_wd001, by = c("LSOA21CD" = "lsoa21cd")) %>%
  left_join(aed_counts, by = c("LSOA21CD" = "lsoa21")) %>%
  mutate(
    aed_count = replace_na(aed_count, 0),
    # Priority score: higher workday density with fewer AEDs = higher priority
    priority_score = workday_density / (aed_count + 1),
    priority_rank = rank(-priority_score, ties.method = "first")
  )

cat("LSOAs with zero 24/7 public AEDs:",
    sum(bh_analysis$aed_count == 0), "of", nrow(bh_analysis), "\n")
```

## Workday Population Analysis

### Map 1: Workday Population Demand with AED Locations

This map shows the Census 2021 workday population density across Brighton & Hove LSOAs. Darker shading indicates higher concentrations of people during the working day. Red points show the locations of existing 24/7 publicly accessible defibrillators.

```{r}
#| label: map-demand
#| fig-width: 10
#| fig-height: 8

tmap_mode("plot")

tm_shape(bh_analysis) +
  tm_polygons(
    fill = "workday_density",
    fill.scale = tm_scale_continuous(
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Workday Pop.\nDensity\n(per km²)"
    ),
    col = "grey60",
    lwd = 0.5
  ) +
  tm_shape(bh_247_sf) +
  tm_symbols(
    fill = "red",
    col = "darkred",
    size = 0.3,
    shape = 21
  ) +
  tm_title("Workday Population Demand & 24/7 Public AED Locations") +
  tm_layout(
    frame = FALSE
  ) +
  tm_scalebar(position = c("left", "bottom"))
```

### Map 2: Interactive Workday Demand & Provision Map

Zoom in to explore individual LSOAs and AED locations. Click on LSOAs for workday population data, or on red markers for AED details.

```{r}
#| label: map-interactive

# Create colour palette for workday density
pal_density <- colorNumeric("Blues", domain = bh_analysis$workday_density)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Choropleth layer
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_density(workday_density),
    fillOpacity = 0.6,
    color = "grey50",
    weight = 1,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Workday density:</em> ",
      format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count, "<br>",
      "<em>Priority rank:</em> ", priority_rank, " of ", nrow(bh_analysis)
    ),
    group = "Workday Density"
  ) %>%
  # AED point markers
  addCircleMarkers(
    data = bh_247_sf,
    radius = 5,
    color = "darkred",
    fillColor = "red",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_density,
    values = bh_analysis$workday_density,
    title = "Workday Density<br>(per km²)"
  ) %>%
  addLayersControl(
    overlayGroups = c("Workday Density", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Map 3: Workday Coverage Gap — Priority Areas for New AEDs

The priority score is calculated as: **workday population density ÷ (number of existing 24/7 public AEDs + 1)**. A higher score indicates an LSOA with high daytime demand but few or no defibrillators — and therefore greater need for new provision.

LSOAs outlined in red are the **top 20 priority** areas. Click on any LSOA for details.

```{r}
#| label: map-priority

# Flag top 20 priority LSOAs
bh_analysis <- bh_analysis %>%
  mutate(top20 = priority_rank <= 20)

# Colour palette for priority score
pal_priority <- colorNumeric("YlOrRd", domain = bh_analysis$priority_score)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # All LSOAs coloured by priority score
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_priority(priority_score),
    fillOpacity = 0.6,
    color = ~ifelse(top20, "red", "grey50"),
    weight = ~ifelse(top20, 3, 1),
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Priority rank:</em> <strong>", priority_rank,
      "</strong> of ", nrow(bh_analysis), "<br>",
      "<em>Priority score:</em> ",
      format(round(priority_score), big.mark = ","), "<br>",
      "<em>Workday density:</em> ",
      format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Priority Score"
  ) %>%
  # Existing AEDs
  addCircleMarkers(
    data = bh_247_sf,
    radius = 4,
    color = "black",
    fillColor = "white",
    fillOpacity = 0.9,
    weight = 1.5,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_priority,
    values = bh_analysis$priority_score,
    title = "Priority Score"
  ) %>%
  addLayersControl(
    overlayGroups = c("Priority Score", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Workday Priority Ranking Table

The table below ranks all Brighton & Hove LSOAs by priority for new 24/7 public defibrillator placement based on **workday population density**.

```{r}
#| label: priority-table

priority_table <- bh_analysis %>%
  st_drop_geometry() %>%
  select(
    Rank = priority_rank,
    LSOA = LSOA21NM,
    `Workday Density (per km²)` = workday_density,
    `24/7 Public AEDs` = aed_count,
    `Priority Score` = priority_score
  ) %>%
  arrange(Rank) %>%
  mutate(
    `Workday Density (per km²)` = format(round(`Workday Density (per km²)`),
                                          big.mark = ","),
    `Priority Score` = format(round(`Priority Score`), big.mark = ",")
  )

# Show top 20
priority_table %>%
  head(20) %>%
  kable(align = c("r", "l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

### Full Workday Ranking — Interactive Map

This map shows all LSOAs coloured by priority rank (darker red = higher priority), with existing 24/7 public AED locations (red markers) and OpenStreetMap telephone box locations (green markers). Telephone boxes represent potential sites for new defibrillator installations.

```{r}
#| label: map-full-ranking

# Colour palette: rank 1 = darkest, rank 165 = lightest
pal_rank <- colorNumeric(
  palette = "YlOrRd",
  domain = bh_analysis$priority_rank,
  reverse = TRUE
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # LSOAs coloured by priority rank
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_rank(priority_rank),
    fillOpacity = 0.5,
    color = "grey50",
    weight = 1,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Priority rank:</em> <strong>", priority_rank,
      "</strong> of ", nrow(bh_analysis), "<br>",
      "<em>Priority score:</em> ",
      format(round(priority_score), big.mark = ","), "<br>",
      "<em>Workday density:</em> ",
      format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Priority Ranking"
  ) %>%
  # Existing 24/7 public AEDs
  addCircleMarkers(
    data = bh_247_sf,
    radius = 5,
    color = "darkred",
    fillColor = "red",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  # OSM telephone boxes
  addCircleMarkers(
    data = phone_sf,
    radius = 5,
    color = "darkgreen",
    fillColor = "#2ca02c",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>Telephone Box</strong><br>",
      "<em>OSM ID:</em> ", id
    ),
    group = "Telephone Boxes (OSM)"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_rank,
    values = bh_analysis$priority_rank,
    title = "Priority Rank",
    labFormat = labelFormat(transform = function(x) round(x))
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("red", "#2ca02c"),
    labels = c("24/7 Public AED", "Telephone Box"),
    title = "Points"
  ) %>%
  addLayersControl(
    overlayGroups = c("Priority Ranking", "24/7 Public AEDs",
                      "Telephone Boxes (OSM)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Full Workday Ranking Table

```{r}
#| label: full-table

priority_table %>%
  kable(align = c("r", "l", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  scroll_box(height = "400px")
```

## Resident Population Analysis

The workday population captures where people are during the day, but cardiac arrests can happen at any time. The **usual resident population** from Census 2021 (table TS001) provides a complementary perspective — reflecting where people live, sleep, and spend evenings and weekends. Comparing the two measures reveals whether AED provision aligns with both daytime and residential demand.

### Build Resident Population Dataset

```{r}
#| label: build-resident-dataset

# Calculate LSOA area in km² using BNG projection for accurate measurement
bh_analysis_bng <- st_transform(bh_analysis, 27700)
bh_analysis$area_km2 <- as.numeric(st_area(bh_analysis_bng)) / 1e6

# Join resident population to the analysis dataset
bh_analysis <- bh_analysis %>%
  left_join(bh_ts001, by = c("LSOA21CD" = "lsoa21cd")) %>%
  mutate(
    resident_density = resident_pop / area_km2,
    # Priority score: higher resident density with fewer AEDs = higher priority
    resident_priority_score = resident_density / (aed_count + 1),
    resident_priority_rank = rank(-resident_priority_score, ties.method = "first")
  )

cat("Resident population density range:",
    format(round(min(bh_analysis$resident_density, na.rm = TRUE)), big.mark = ","),
    "to",
    format(round(max(bh_analysis$resident_density, na.rm = TRUE)), big.mark = ","),
    "persons per km²\n")

cat("LSOAs with zero 24/7 public AEDs:",
    sum(bh_analysis$aed_count == 0), "of", nrow(bh_analysis), "\n")
```

### Map 4: Resident Population Density with AED Locations

This map shows the Census 2021 usual resident population density across Brighton & Hove LSOAs. Darker shading indicates more densely populated residential areas. Red points show the locations of existing 24/7 publicly accessible defibrillators.

```{r}
#| label: map-resident-demand
#| fig-width: 10
#| fig-height: 8

tmap_mode("plot")

tm_shape(bh_analysis) +
  tm_polygons(
    fill = "resident_density",
    fill.scale = tm_scale_continuous(
      values = "brewer.purples"
    ),
    fill.legend = tm_legend(
      title = "Resident Pop.\nDensity\n(per km²)"
    ),
    col = "grey60",
    lwd = 0.5
  ) +
  tm_shape(bh_247_sf) +
  tm_symbols(
    fill = "red",
    col = "darkred",
    size = 0.3,
    shape = 21
  ) +
  tm_title("Resident Population Density & 24/7 Public AED Locations") +
  tm_layout(
    frame = FALSE
  ) +
  tm_scalebar(position = c("left", "bottom"))
```

### Map 5: Interactive Resident Population & AED Provision Map

Zoom in to explore resident population density by LSOA. Click on LSOAs for population data, or on red markers for AED details.

```{r}
#| label: map-resident-interactive

# Create colour palette for resident density
pal_resident <- colorNumeric("Purples",
                              domain = bh_analysis$resident_density)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_resident(resident_density),
    fillOpacity = 0.6,
    color = "grey50",
    weight = 1,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Usual residents:</em> ",
      format(resident_pop, big.mark = ","), "<br>",
      "<em>Resident density:</em> ",
      format(round(resident_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count, "<br>",
      "<em>Resident priority rank:</em> ",
      resident_priority_rank, " of ", nrow(bh_analysis)
    ),
    group = "Resident Density"
  ) %>%
  addCircleMarkers(
    data = bh_247_sf,
    radius = 5,
    color = "darkred",
    fillColor = "red",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_resident,
    values = bh_analysis$resident_density,
    title = "Resident Density<br>(per km²)"
  ) %>%
  addLayersControl(
    overlayGroups = c("Resident Density", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Map 6: Resident Population Coverage Gap — Priority Areas

The resident priority score is calculated as: **resident population density ÷ (number of existing 24/7 public AEDs + 1)**. LSOAs outlined in purple are the **top 20 priority** areas based on resident population demand.

```{r}
#| label: map-resident-priority

bh_analysis <- bh_analysis %>%
  mutate(resident_top20 = resident_priority_rank <= 20)

# Colour palette for resident priority score
pal_res_priority <- colorNumeric("YlOrRd",
                                  domain = bh_analysis$resident_priority_score)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_res_priority(resident_priority_score),
    fillOpacity = 0.6,
    color = ~ifelse(resident_top20, "purple", "grey50"),
    weight = ~ifelse(resident_top20, 3, 1),
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Resident priority rank:</em> <strong>",
      resident_priority_rank,
      "</strong> of ", nrow(bh_analysis), "<br>",
      "<em>Resident priority score:</em> ",
      format(round(resident_priority_score), big.mark = ","), "<br>",
      "<em>Resident density:</em> ",
      format(round(resident_density), big.mark = ","), " per km²<br>",
      "<em>Usual residents:</em> ",
      format(resident_pop, big.mark = ","), "<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Resident Priority"
  ) %>%
  addCircleMarkers(
    data = bh_247_sf,
    radius = 4,
    color = "black",
    fillColor = "white",
    fillOpacity = 0.9,
    weight = 1.5,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_res_priority,
    values = bh_analysis$resident_priority_score,
    title = "Resident Priority<br>Score"
  ) %>%
  addLayersControl(
    overlayGroups = c("Resident Priority", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Resident Population Priority Ranking Table

The table below ranks all Brighton & Hove LSOAs by priority for new AED placement based on **usual resident population density**.

```{r}
#| label: resident-priority-table

resident_priority_table <- bh_analysis %>%
  st_drop_geometry() %>%
  select(
    Rank = resident_priority_rank,
    LSOA = LSOA21NM,
    `Usual Residents` = resident_pop,
    `Resident Density (per km²)` = resident_density,
    `24/7 Public AEDs` = aed_count,
    `Priority Score` = resident_priority_score
  ) %>%
  arrange(Rank) %>%
  mutate(
    `Usual Residents` = format(`Usual Residents`, big.mark = ","),
    `Resident Density (per km²)` = format(round(`Resident Density (per km²)`),
                                            big.mark = ","),
    `Priority Score` = format(round(`Priority Score`), big.mark = ",")
  )

# Show top 20
resident_priority_table %>%
  head(20) %>%
  kable(align = c("r", "l", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

### Full Resident Ranking Table

```{r}
#| label: full-resident-table

resident_priority_table %>%
  kable(align = c("r", "l", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  scroll_box(height = "400px")
```

### Full Resident Ranking — Interactive Map

This map shows all LSOAs coloured by resident population priority rank (darker purple = higher priority), with existing 24/7 public AED locations (red markers) and OpenStreetMap telephone box locations (green markers).

```{r}
#| label: map-resident-full-ranking

# Colour palette: rank 1 = darkest, highest rank = lightest
pal_res_rank <- colorNumeric(
  palette = "Purples",
  domain = bh_analysis$resident_priority_rank,
  reverse = TRUE
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_res_rank(resident_priority_rank),
    fillOpacity = 0.5,
    color = "grey50",
    weight = 1,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Resident priority rank:</em> <strong>",
      resident_priority_rank,
      "</strong> of ", nrow(bh_analysis), "<br>",
      "<em>Resident priority score:</em> ",
      format(round(resident_priority_score), big.mark = ","), "<br>",
      "<em>Resident density:</em> ",
      format(round(resident_density), big.mark = ","), " per km²<br>",
      "<em>Usual residents:</em> ",
      format(resident_pop, big.mark = ","), "<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Resident Priority Ranking"
  ) %>%
  addCircleMarkers(
    data = bh_247_sf,
    radius = 5,
    color = "darkred",
    fillColor = "red",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addCircleMarkers(
    data = phone_sf,
    radius = 5,
    color = "darkgreen",
    fillColor = "#2ca02c",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>Telephone Box</strong><br>",
      "<em>OSM ID:</em> ", id
    ),
    group = "Telephone Boxes (OSM)"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_res_rank,
    values = bh_analysis$resident_priority_rank,
    title = "Resident<br>Priority Rank",
    labFormat = labelFormat(transform = function(x) round(x))
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("red", "#2ca02c"),
    labels = c("24/7 Public AED", "Telephone Box"),
    title = "Points"
  ) %>%
  addLayersControl(
    overlayGroups = c("Resident Priority Ranking", "24/7 Public AEDs",
                      "Telephone Boxes (OSM)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

## Comparing Workday and Resident Population Demand

Different population measures can lead to different conclusions about where AEDs are most needed. The workday population reflects daytime demand (offices, shops, transport hubs), while the resident population reflects overnight, evening, and weekend demand (homes, residential streets). Comparing the two reveals which areas are consistently high-priority and which shift depending on the measure used.

### Density Comparison

```{r}
#| label: density-scatter
#| fig-width: 8
#| fig-height: 7

comparison_df <- bh_analysis %>%
  st_drop_geometry() %>%
  select(LSOA21NM, workday_density, resident_density, aed_count,
         priority_rank, resident_priority_rank)

ggplot(comparison_df,
       aes(x = workday_density, y = resident_density)) +
  geom_point(aes(colour = factor(pmin(aed_count, 3))),
             size = 2.5, alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", colour = "grey40") +
  scale_colour_manual(
    values = c("0" = "#d73027", "1" = "#fee08b",
               "2" = "#91cf60", "3" = "#1a9850"),
    labels = c("0" = "0", "1" = "1", "2" = "2", "3" = "3+"),
    name = "24/7 Public\nAEDs"
  ) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Workday vs. Resident Population Density by LSOA",
    subtitle = "Points above the diagonal have higher resident than workday density",
    x = "Workday Population Density (per km²)",
    y = "Resident Population Density (per km²)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

LSOAs above the dashed line have higher resident density than workday density — predominantly residential areas. LSOAs below the line attract more people during the day than they house — commercial, retail, or transport hubs. Red points (no AEDs) in either high-density zone indicate gaps in coverage.

### Priority Rank Differences

The table below shows LSOAs where the priority ranking changes most between the two measures. A large positive rank difference means the LSOA is ranked as much higher priority by the **resident** analysis than the workday analysis (i.e. it is a densely populated residential area that is less busy during the day). A large negative difference means the reverse — a daytime hotspot that is less densely populated residentially.

```{r}
#| label: rank-change-table

rank_changes <- bh_analysis %>%
  st_drop_geometry() %>%
  mutate(
    rank_diff = priority_rank - resident_priority_rank
  ) %>%
  select(
    LSOA = LSOA21NM,
    `Workday Rank` = priority_rank,
    `Resident Rank` = resident_priority_rank,
    `Rank Difference` = rank_diff,
    `Workday Density` = workday_density,
    `Resident Density` = resident_density,
    `24/7 Public AEDs` = aed_count
  ) %>%
  arrange(desc(abs(`Rank Difference`)))

# Show top 20 largest rank changes
rank_changes %>%
  head(20) %>%
  mutate(
    `Workday Density` = format(round(`Workday Density`), big.mark = ","),
    `Resident Density` = format(round(`Resident Density`), big.mark = ",")
  ) %>%
  kable(align = c("l", "r", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

### Map 7: Priority Rank Differences

This map visualises how priority rankings shift between the two population measures. **Blue** LSOAs are ranked as higher priority under the **workday** analysis (daytime hotspots); **red** LSOAs are ranked as higher priority under the **resident** analysis (residential demand). Neutral colours indicate consistent priority under both measures.

```{r}
#| label: map-rank-diff

bh_analysis <- bh_analysis %>%
  mutate(rank_diff = priority_rank - resident_priority_rank)

# Diverging palette: blue = workday priority higher, red = resident priority higher
max_abs_diff <- max(abs(bh_analysis$rank_diff), na.rm = TRUE)
pal_diff <- colorNumeric(
  palette = "RdBu",
  domain = c(-max_abs_diff, max_abs_diff),
  reverse = TRUE
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_diff(rank_diff),
    fillOpacity = 0.6,
    color = "grey50",
    weight = 1,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Workday priority rank:</em> ", priority_rank, "<br>",
      "<em>Resident priority rank:</em> ", resident_priority_rank, "<br>",
      "<em>Rank difference:</em> <strong>",
      ifelse(rank_diff > 0, paste0("+", rank_diff), rank_diff),
      "</strong><br>",
      "<hr style='margin:4px 0'>",
      "<em>Workday density:</em> ",
      format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>Resident density:</em> ",
      format(round(resident_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Rank Difference"
  ) %>%
  addCircleMarkers(
    data = bh_247_sf,
    radius = 4,
    color = "black",
    fillColor = "white",
    fillOpacity = 0.9,
    weight = 1.5,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_diff,
    values = bh_analysis$rank_diff,
    title = "Rank Difference<br>(Workday − Resident)",
    labFormat = labelFormat(
      transform = function(x) round(x)
    )
  ) %>%
  addLayersControl(
    overlayGroups = c("Rank Difference", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Map 8: Combined Workday & Resident Priority Ranking

This map combines both population measures into a single priority ranking by averaging the workday and resident priority ranks. This gives equal weight to daytime and residential demand, producing a balanced view of where new AEDs are most needed across all times of day. Darker shading indicates higher combined priority. LSOAs outlined in red are the **top 20** under the combined ranking.

```{r}
#| label: map-combined-ranking

# Compute combined priority: average of the two ranks (lower = higher priority)
bh_analysis <- bh_analysis %>%
  mutate(
    combined_avg_rank = (priority_rank + resident_priority_rank) / 2,
    combined_priority_rank = rank(combined_avg_rank, ties.method = "first"),
    combined_top20 = combined_priority_rank <= 20
  )

# Colour palette: rank 1 = darkest
pal_combined <- colorNumeric(
  palette = "YlOrRd",
  domain = bh_analysis$combined_priority_rank,
  reverse = TRUE
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    data = bh_analysis,
    fillColor = ~pal_combined(combined_priority_rank),
    fillOpacity = 0.5,
    color = ~ifelse(combined_top20, "red", "grey50"),
    weight = ~ifelse(combined_top20, 3, 1),
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Combined priority rank:</em> <strong>",
      combined_priority_rank,
      "</strong> of ", nrow(bh_analysis), "<br>",
      "<em>Workday rank:</em> ", priority_rank,
      " | <em>Resident rank:</em> ", resident_priority_rank, "<br>",
      "<hr style='margin:4px 0'>",
      "<em>Workday density:</em> ",
      format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>Resident density:</em> ",
      format(round(resident_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Combined Priority"
  ) %>%
  addCircleMarkers(
    data = bh_247_sf,
    radius = 5,
    color = "darkred",
    fillColor = "red",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addCircleMarkers(
    data = phone_sf,
    radius = 5,
    color = "darkgreen",
    fillColor = "#2ca02c",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>Telephone Box</strong><br>",
      "<em>OSM ID:</em> ", id
    ),
    group = "Telephone Boxes (OSM)"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal_combined,
    values = bh_analysis$combined_priority_rank,
    title = "Combined<br>Priority Rank",
    labFormat = labelFormat(transform = function(x) round(x))
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("red", "#2ca02c"),
    labels = c("24/7 Public AED", "Telephone Box"),
    title = "Points"
  ) %>%
  addLayersControl(
    overlayGroups = c("Combined Priority", "24/7 Public AEDs",
                      "Telephone Boxes (OSM)"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Combined Priority Ranking Table

```{r}
#| label: combined-priority-table

combined_table <- bh_analysis %>%
  st_drop_geometry() %>%
  select(
    `Combined Rank` = combined_priority_rank,
    LSOA = LSOA21NM,
    `Workday Rank` = priority_rank,
    `Resident Rank` = resident_priority_rank,
    `Workday Density` = workday_density,
    `Resident Density` = resident_density,
    `24/7 Public AEDs` = aed_count
  ) %>%
  arrange(`Combined Rank`) %>%
  mutate(
    `Workday Density` = format(round(`Workday Density`), big.mark = ","),
    `Resident Density` = format(round(`Resident Density`), big.mark = ",")
  )

combined_table %>%
  head(20) %>%
  kable(align = c("r", "l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

### Consistently High-Priority LSOAs

LSOAs that rank in the **top 20 under both** workday and resident population analyses represent the most robust candidates for new AED placement — they have high demand regardless of which population measure is used.

```{r}
#| label: consistent-priority

consistent_top <- bh_analysis %>%
  st_drop_geometry() %>%
  filter(priority_rank <= 20, resident_priority_rank <= 20) %>%
  select(
    LSOA = LSOA21NM,
    `Workday Rank` = priority_rank,
    `Resident Rank` = resident_priority_rank,
    `Workday Density` = workday_density,
    `Resident Density` = resident_density,
    `24/7 Public AEDs` = aed_count
  ) %>%
  arrange(`Workday Rank`) %>%
  mutate(
    `Workday Density` = format(round(`Workday Density`), big.mark = ","),
    `Resident Density` = format(round(`Resident Density`), big.mark = ",")
  )

n_consistent <- nrow(consistent_top)

cat(n_consistent, "LSOAs appear in the top 20 under both measures:\n\n")

consistent_top %>%
  kable(align = c("l", "r", "r", "r", "r", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

### Map 9: Top 15 Consistently High-Priority LSOAs

The map below highlights the top 15 LSOAs from the combined priority ranking that also appear in the top 20 under **both** the workday and resident analyses. These are grouped into three tiers to help identify the most urgent candidates for new AED provision.


```{r}
library(leaflet)
library(dplyr)

# 1. DATA PREPARATION
# Ensure top15 is created correctly in the current session
bh_analysis <- bh_analysis %>%
  mutate(consistent = priority_rank <= 20 & resident_priority_rank <= 20)

top15 <- bh_analysis %>%
  filter(consistent) %>%
  arrange(combined_priority_rank) %>%
  head(15) %>%
  mutate(
    tier = case_when(
      row_number() <= 5  ~ "Top 5",
      row_number() <= 10 ~ "6th–10th",
      TRUE               ~ "11th–15th"
    ),
    tier = factor(tier, levels = c("Top 5", "6th–10th", "11th–15th"))
  )

bg_lsoas <- bh_analysis %>% filter(!LSOA21CD %in% top15$LSOA21CD)

# 2. COLOR PALETTE
# Using ColorBrewer-style "YlOrRd" logic for a clear heat fade
pal <- colorFactor(
  palette = c("#d7191c", "#fdae61", "#ffffbf"), 
  levels = c("Top 5", "6th–10th", "11th–15th")
)

# 3. MAP GENERATION
leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # Background: All other LSOAs
  addPolygons(
    data = bg_lsoas,
    fillColor = "#e0e0e0",
    fillOpacity = 0.3,
    color = "grey70",
    weight = 0.5,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Combined rank:</em> ", combined_priority_rank,
      " of ", nrow(bh_analysis)
    ),
    group = "Other LSOAs"
  ) %>%
  
  # Foreground: Top 15 Priority LSOAs
  addPolygons(
    data = top15,
    fillColor = ~pal(tier),
    fillOpacity = 0.7,
    color = "black",
    weight = 2,
    popup = ~paste0(
      "<strong>", LSOA21NM, "</strong><br>",
      "<em>Tier:</em> <strong>", tier, "</strong><br>",
      "<em>Combined rank:</em> ", combined_priority_rank, "<br>",
      "<em>Workday rank:</em> ", priority_rank,
      " | <em>Resident rank:</em> ", resident_priority_rank, "<br>",
      "<hr style='margin:4px 0'>",
      "<em>Workday density:</em> ", format(round(workday_density), big.mark = ","), " per km²<br>",
      "<em>Resident density:</em> ", format(round(resident_density), big.mark = ","), " per km²<br>",
      "<em>24/7 Public AEDs:</em> ", aed_count
    ),
    group = "Top 15 Priority LSOAs"
  ) %>%
  
  # Points: 24/7 Public AEDs
  addCircleMarkers(
    data = bh_247_sf,
    radius = 4,
    color = "black",
    fillColor = "white",
    fillOpacity = 0.9,
    weight = 1.5,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code
    ),
    group = "24/7 Public AEDs"
  ) %>%
  addCircleMarkers(
    data = phone_sf,
    radius = 5,
    color = "darkgreen",
    fillColor = "#2ca02c",
    fillOpacity = 0.8,
    weight = 1,
    popup = ~paste0(
      "<strong>Telephone Box</strong><br>",
      "<em>OSM ID:</em> ", id
    ),
    group = "Telephone Boxes (OSM)"
  ) %>%
  
  # Legend and Controls
  addLegend(
    position = "bottomright",
    pal = pal,
    values = top15$tier,
    title = "Priority Tier",
    opacity = 0.7
  ) %>%
  addLayersControl(
    overlayGroups = c("Top 15 Priority LSOAs", "Other LSOAs", "24/7 Public AEDs"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

## Summary

```{r}
#| label: summary-stats
#| echo: false

n_total <- nrow(bh_analysis)
n_zero <- sum(bh_analysis$aed_count == 0)
n_aeds <- nrow(bh_247_sf)
top1_wd <- bh_analysis %>% filter(priority_rank == 1) %>% st_drop_geometry()
top1_res <- bh_analysis %>% filter(resident_priority_rank == 1) %>% st_drop_geometry()
n_consistent <- bh_analysis %>%
  st_drop_geometry() %>%
  filter(priority_rank <= 20, resident_priority_rank <= 20) %>%
  nrow()
total_residents <- sum(bh_analysis$resident_pop, na.rm = TRUE)
```

**Key findings:**

-   Brighton & Hove has **`r n_total` LSOAs**, **`r format(total_residents, big.mark = ",")` usual residents**, and **`r n_aeds` 24/7 publicly accessible AEDs**
-   **`r n_zero` LSOAs (`r round(100 * n_zero / n_total)`%)** have **no** 24/7 public AED
-   **Workday analysis:** The highest-priority LSOA is **`r top1_wd$LSOA21NM`** with a workday density of `r format(round(top1_wd$workday_density), big.mark = ",")` persons per km² and `r top1_wd$aed_count` existing 24/7 public AED(s)
-   **Resident analysis:** The highest-priority LSOA is **`r top1_res$LSOA21NM`** with a resident density of `r format(round(top1_res$resident_density), big.mark = ",")` persons per km² and `r top1_res$aed_count` existing 24/7 public AED(s)
-   **`r n_consistent` LSOAs** appear in the top 20 priority under **both** workday and resident population measures — these are the strongest candidates for new AED provision
-   The priority ranking combines both demand (population density) and existing supply, meaning LSOAs that already have good coverage are ranked lower regardless of their density
-   Areas where the two rankings diverge most highlight the difference between daytime commercial centres and densely populated residential neighbourhoods

------------------------------------------------------------------------

*Data sourced from [The Circuit](https://www.defibfinder.uk/), [ONS Census 2021 Workday Population](https://www.nomisweb.co.uk/sources/census_2021_wd), and [ONS Census 2021 Usual Resident Population](https://www.nomisweb.co.uk/sources/census_2021_bulk). Is this an emergency? This website is used to locate defibrillators, but it is not intended for use in an emergency. If you require urgent medical assistance, call 999 now.*
