---
title: "UK Defibrillator Locations"
author: "Adam Dennett"
format: html
execute:
  warning: false
  message: false
---

## Introduction

This document maps the locations of Automated External Defibrillators (AEDs) across the UK. The data is sourced from [The Circuit](https://www.defibfinder.uk/) and contains over 116,000 defibrillator records.

Points are coloured by **availability** (blue = 24/7, orange = Varied) and shaded by **access type** (darker = Public, lighter = Restricted).

## Load Libraries and Data

```{r}
#| label: load-data
#| message: false
#| warning: false

library(readxl)
library(sf)
library(dplyr)
library(leaflet)

# ---- Basemap, no API key ---------------------------------------------
# CARTO require a key on their raster tiles, and without one the tiles
# carry an "API key required" watermark. That failed in the only way
# that really matters: silently. Any render from a shell without
# CARTO_KEY in its environment produced watermarked maps and no other
# symptom, so the fault was found by eye or not at all.
#
# So the tiles come from Esri's World Gray Canvas instead, which needs
# no key. Of the keyless alternatives it is the one that actually is
# keyless: Stadia's tiles return 401 without a key and Jawg's return
# 400, both having moved to requiring one.
#
# This still wraps addProviderTiles() so that no call site has to
# change. A CartoDB.* provider is served from Esri; anything else is
# passed straight through to leaflet. Esri splits the canvas and the
# place names into two services where CARTO bakes them into one, so the
# Positron-style variants map onto one or both of them, and the
# DarkMatter ones onto Esri's dark canvas.
#
# maxNativeZoom is set because Esri publishes these canvases to zoom 16
# and above that the server returns a grey tile reading "Map data not
# yet available" rather than a 404 - a silent failure again. Leaflet
# upsamples level 16 instead of asking for 17, so the basemap softens
# past that point and the data drawn over it stays sharp.
addProviderTiles <- function(map, provider, ...) {
  nm <- as.character(provider)[1]
  if (!startsWith(nm, "CartoDB."))
    return(leaflet::addProviderTiles(map, provider, ...))
  nm <- substring(nm, 9L)

  dark <- startsWith(nm, "DarkMatter")
  shade <- if (dark) "Dark" else "Light"
  esri <- function(part) sprintf(paste0(
    "https://server.arcgisonline.com/ArcGIS/rest/services/Canvas/",
    "World_%s_Gray_%s/MapServer/tile/{z}/{y}/{x}"), shade, part)

  want_base   <- !grepl("OnlyLabels$", nm)
  want_labels <- !grepl("NoLabels$", nm)

  dots <- list(...)
  opts <- leaflet::tileOptions(maxNativeZoom = 16, maxZoom = 20)
  if (!is.null(dots$options)) {
    opts <- utils::modifyList(opts, dots$options); dots$options <- NULL
  }
  attrib <- paste0(
    'Tiles &copy; <a href="https://www.esri.com/">Esri</a> &mdash; Esri, ',
    'HERE, Garmin, &copy; ',
    '<a href="https://www.openstreetmap.org/copyright">OpenStreetMap</a>',
    ' contributors')

  if (want_base)
    map <- do.call(leaflet::addTiles, c(list(
      map, urlTemplate = esri("Base"), attribution = attrib,
      options = opts), dots))
  # Only one of the two layers carries the credit, or every map would
  # print it twice in the corner.
  if (want_labels)
    map <- do.call(leaflet::addTiles, c(list(
      map, urlTemplate = esri("Reference"),
      attribution = if (want_base) "" else attrib,
      options = opts), dots))
  map
}
# ----------------------------------------------------------------------
library(htmltools)

defib <- read_excel("defibrillator_data.xlsx",
                     sheet = "data_extract_2026-02-02")

glimpse(defib)
```

## Clean and Convert to Spatial Data

```{r}
#| label: spatial-convert

# Remove rows with missing coordinates
defib_clean <- defib %>%
  filter(!is.na(lat), !is.na(long))

# Convert to sf object
defib_sf <- st_as_sf(defib_clean, coords = c("long", "lat"), crs = 4326)
```

## Assign Colours

```{r}
#| label: colour-mapping

# Colour scheme:
#   24/7 Access + Public     -> dark blue   (#08519c)
#   24/7 Access + Restricted -> light blue  (#6baed6)
#   Varied Access + Public   -> dark orange (#d94701)
#   Varied Access + Restricted -> light orange (#fdae6b)

defib_sf <- defib_sf %>%
  mutate(
    point_colour = case_when(
      defibrillators_availability == "24/7 Access" &
        defibrillators_access_type == "Public"     ~ "#08519c",
      defibrillators_availability == "24/7 Access" &
        defibrillators_access_type == "Restricted" ~ "#6baed6",
      defibrillators_availability == "Varied Access" &
        defibrillators_access_type == "Public"     ~ "#d94701",
      defibrillators_availability == "Varied Access" &
        defibrillators_access_type == "Restricted" ~ "#fdae6b",
      TRUE ~ "#999999"
    )
  )

# Summary of colour categories
defib_sf %>%
  st_drop_geometry() %>%
  count(defibrillators_availability, defibrillators_access_type, point_colour,
        name = "count") %>%
  arrange(desc(count))
```

## Interactive Map

```{r}
#| label: leaflet-map

leaflet(defib_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    radius = 3,
    color = ~point_colour,
    stroke = FALSE,
    fillOpacity = 0.6,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code, "<br>",
      "<em>Availability:</em> ", defibrillators_availability, "<br>",
      "<em>Access:</em> ", defibrillators_access_type
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("#08519c", "#6baed6", "#d94701", "#fdae6b"),
    labels = c("24/7 — Public", "24/7 — Restricted",
               "Varied — Public", "Varied — Restricted"),
    title = "Availability — Access",
    opacity = 0.8
  )
```
