---
title: "UK Defibrillator Locations"
author: "Adam Dennett"
format: html
execute:
  warning: false
  message: false
---

## Introduction

This document maps the locations of Automated External Defibrillators (AEDs) across the UK. The data is sourced from [The Circuit](https://www.defibfinder.uk/) and contains over 116,000 defibrillator records.

Points are coloured by **availability** (blue = 24/7, orange = Varied) and shaded by **access type** (darker = Public, lighter = Restricted).

## Load Libraries and Data

```{r}
#| label: load-data
#| message: false
#| warning: false

library(readxl)
library(sf)
library(dplyr)
library(leaflet)

# ---- CARTO basemap API key -------------------------------------------
# CARTO now require a key on their raster tiles; without one the tiles
# carry an "API key required" watermark. leaflet-providers has no slot
# for a key, so this wraps addProviderTiles(): CartoDB providers are
# rebuilt as a keyed addTiles() call and everything else is passed
# straight through to leaflet. Attribution is set explicitly, because
# addTiles() does not inherit the provider's and keeping the CARTO and
# OpenStreetMap credits visible is a condition of the free tier.
# The key is read from CARTO_KEY in ~/.Renviron. If it is unset the maps
# still render, just watermarked, so clones and CI are unaffected.
addProviderTiles <- function(map, provider, ...) {
  variants <- c(
    Positron             = "light_all",
    PositronNoLabels     = "light_nolabels",
    PositronOnlyLabels   = "light_only_labels",
    DarkMatter           = "dark_all",
    DarkMatterNoLabels   = "dark_nolabels",
    DarkMatterOnlyLabels = "dark_only_labels",
    Voyager              = "rastertiles/voyager",
    VoyagerNoLabels      = "rastertiles/voyager_nolabels",
    VoyagerOnlyLabels    = "rastertiles/voyager_only_labels",
    VoyagerLabelsUnder   = "rastertiles/voyager_labels_under")
  key <- Sys.getenv("CARTO_KEY", "")
  nm  <- as.character(provider)[1]
  if (startsWith(nm, "CartoDB.")) nm <- substring(nm, 9L)
  if (!nzchar(key) || !nm %in% names(variants))
    return(leaflet::addProviderTiles(map, provider, ...))
  dots <- list(...)
  opts <- leaflet::tileOptions(subdomains = "abcd", maxZoom = 20)
  if (!is.null(dots$options)) {
    opts <- utils::modifyList(opts, dots$options); dots$options <- NULL
  }
  do.call(leaflet::addTiles, c(list(
    map,
    urlTemplate = sprintf(
      "https://{s}.basemaps.cartocdn.com/%s/{z}/{x}/{y}{r}.png?key=%s",
      variants[[nm]], key),
    attribution = paste0(
      '&copy; <a href="https://www.openstreetmap.org/copyright">OpenStreetMap</a>',
      ' contributors &copy; <a href="https://carto.com/attributions">CARTO</a>'),
    options = opts), dots))
}
# ----------------------------------------------------------------------
library(htmltools)

defib <- read_excel("defibrillator_data.xlsx",
                     sheet = "data_extract_2026-02-02")

glimpse(defib)
```

## Clean and Convert to Spatial Data

```{r}
#| label: spatial-convert

# Remove rows with missing coordinates
defib_clean <- defib %>%
  filter(!is.na(lat), !is.na(long))

# Convert to sf object
defib_sf <- st_as_sf(defib_clean, coords = c("long", "lat"), crs = 4326)
```

## Assign Colours

```{r}
#| label: colour-mapping

# Colour scheme:
#   24/7 Access + Public     -> dark blue   (#08519c)
#   24/7 Access + Restricted -> light blue  (#6baed6)
#   Varied Access + Public   -> dark orange (#d94701)
#   Varied Access + Restricted -> light orange (#fdae6b)

defib_sf <- defib_sf %>%
  mutate(
    point_colour = case_when(
      defibrillators_availability == "24/7 Access" &
        defibrillators_access_type == "Public"     ~ "#08519c",
      defibrillators_availability == "24/7 Access" &
        defibrillators_access_type == "Restricted" ~ "#6baed6",
      defibrillators_availability == "Varied Access" &
        defibrillators_access_type == "Public"     ~ "#d94701",
      defibrillators_availability == "Varied Access" &
        defibrillators_access_type == "Restricted" ~ "#fdae6b",
      TRUE ~ "#999999"
    )
  )

# Summary of colour categories
defib_sf %>%
  st_drop_geometry() %>%
  count(defibrillators_availability, defibrillators_access_type, point_colour,
        name = "count") %>%
  arrange(desc(count))
```

## Interactive Map

```{r}
#| label: leaflet-map

leaflet(defib_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    radius = 3,
    color = ~point_colour,
    stroke = FALSE,
    fillOpacity = 0.6,
    popup = ~paste0(
      "<strong>", location_name, "</strong><br>",
      address_line1, "<br>",
      address_city, " ", address_post_code, "<br>",
      "<em>Availability:</em> ", defibrillators_availability, "<br>",
      "<em>Access:</em> ", defibrillators_access_type
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("#08519c", "#6baed6", "#d94701", "#fdae6b"),
    labels = c("24/7 — Public", "24/7 — Restricted",
               "Varied — Public", "Varied — Restricted"),
    title = "Availability — Access",
    opacity = 0.8
  )
```
